---
title: "GroupA_HW1"
author: "S. Cattonar, L.Ricatti, M. Rizwan, D. Rosa, A. Valle"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
date: "2024-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CS - Chapter 1 

### Ex 1.1
(Luca)
*Exponential random variable, X $\geq$ 0, has p.d.f. f(x) = $\lambda \exp(-\lambda x)$.*    
  
*1. Find the c.d.f. and the quantile function for X.*  
*2. Find Pr(X < $\lambda$) and the median of X.*  
*3. Find the mean and variance of X.*  
\
**Solution**

1. C.D.F. and Quantile Function:

   The cumulative distribution function (c.d.f.) F(x) is:
   $$ F(x) = \int_0^x \lambda e^{-\lambda t} dt = 1 - e^{-\lambda x}, \quad x \geq 0 $$

   The quantile function Q(p) is the inverse of F(x):
   $$ Q(p) = -\frac{1}{\lambda} \ln(1-p), \quad 0 \leq p < 1 $$

2. Pr(X < \(\lambda\)) and Median:

   Pr(X < \(\lambda\)) = F(\(\lambda\)) = \(1 - e^{-\lambda\lambda} = 1 - e^{-1} \approx 0.6321\)

   For the median, we solve F(x) = 0.5:
   $$ 1 - e^{-\lambda x} = 0.5 $$
   $$ x = -\frac{1}{\lambda} \ln(0.5) = \frac{\ln(2)}{\lambda} $$

3. Mean and Variance:

   Mean: E[X] = \(\int_0^\infty x \lambda e^{-\lambda x} dx = \frac{1}{\lambda}\)

   Variance: Var(X) = E[X²] - (E[X])² = \(\frac{2}{\lambda^2} - (\frac{1}{\lambda})^2 = \frac{1}{\lambda^2}\)

**Comments on the solution:**

1. The exponential distribution is characterized by its rate parameter \(\lambda\), which determines both its shape and scale.

2. Interestingly, Pr(X < \(\lambda\)) is always approximately 0.6321, regardless of the value of \(\lambda\). This is a unique property of the exponential distribution.

3. The median of the distribution is \(\frac{\ln(2)}{\lambda}\), which is always less than the mean (\(\frac{1}{\lambda}\)) due to the distribution's right-skewness.

4. The mean and variance are both functions of \(\lambda\). As \(\lambda\) increases, both the mean and variance decrease, indicating that larger values of \(\lambda\) result in the distribution being more concentrated near zero.

5. The standard deviation of the distribution is equal to its mean, which is a distinctive feature of the exponential distribution.



### Ex 1.6
(Luca)
*Let X and Y be non-independent random variables, such that var(X) = $\sigma_x^2$, var(Y) = $\sigma_y^2$ and cov(X,Y) = $\sigma_{xy}^2$. Using the result from Section 1.6.2, find var(X + Y) and var(X - Y).*\
\
**Solution**

Using the formula for linear transformations of random vectors from Section 1.6.2:

$\mathbf{AX} \sim N(\mathbf{A}\boldsymbol{\mu}, \mathbf{A\Sigma A^T})$

1. Define a random vector and its covariance matrix:

   $\mathbf{X} = \begin{pmatrix} X \\ Y \end{pmatrix}$

   $\boldsymbol{\Sigma} = \begin{pmatrix} \sigma_x^2 & \sigma_{xy}^2 \\ \sigma_{xy}^2 & \sigma_y^2 \end{pmatrix}$

2. For var(X + Y):
   Let $\mathbf{A} = \begin{pmatrix} 1 & 1 \end{pmatrix}$

   $\text{var}(X + Y) = \mathbf{A\Sigma A^T} = \begin{pmatrix} 1 & 1 \end{pmatrix} \begin{pmatrix} \sigma_x^2 & \sigma_{xy}^2 \\ \sigma_{xy}^2 & \sigma_y^2 \end{pmatrix} \begin{pmatrix} 1 \\ 1 \end{pmatrix}$

   $= \begin{pmatrix} 1 & 1 \end{pmatrix} \begin{pmatrix} \sigma_x^2 + \sigma_{xy}^2 \\ \sigma_{xy}^2 + \sigma_y^2 \end{pmatrix} = \sigma_x^2 + 2\sigma_{xy}^2 + \sigma_y^2$

3. For var(X - Y):
   Let $\mathbf{A} = \begin{pmatrix} 1 & -1 \end{pmatrix}$

   $\text{var}(X - Y) = \mathbf{A\Sigma A^T} = \begin{pmatrix} 1 & -1 \end{pmatrix} \begin{pmatrix} \sigma_x^2 & \sigma_{xy}^2 \\ \sigma_{xy}^2 & \sigma_y^2 \end{pmatrix} \begin{pmatrix} 1 \\ -1 \end{pmatrix}$

   $= \begin{pmatrix} 1 & -1 \end{pmatrix} \begin{pmatrix} \sigma_x^2 - \sigma_{xy}^2 \\ -(\sigma_{xy}^2 - \sigma_y^2) \end{pmatrix} = \sigma_x^2 - 2\sigma_{xy}^2 + \sigma_y^2$

Therefore:\
\
$\text{var}(X + Y) = \sigma_x^2 + 2\sigma_{xy}^2 + \sigma_y^2$\
\
\
$\text{var}(X - Y) = \sigma_x^2 - 2\sigma_{xy}^2 + \sigma_y^2$\
\
**Comments on the solution:**

1. For the sum (X + Y), the covariance term is added twice, potentially increasing the overall variance if X and Y are positively correlated.

2. For the difference (X - Y), the covariance term is subtracted twice, potentially decreasing the overall variance if X and Y are positively correlated.

3. If X and Y are independent (i.e., $\sigma_{xy}^2 = 0$), the variances of their sum and difference would both simplify to $\sigma_x^2 + \sigma_y^2$.


### Ex 3.5
*Consider solving the matrix equation Ax = y for x, where y is a known n vector and A is a known n × n matrix...*

**Solution**



### Ex 3.6
*The empirical cumulative distribution function for a set of measurements {xi : i = 1, . . . n} is ...*
**Solution**



## FSDS - Chapter 2

### Ex 2.8
(Luca)
*Each time a person shops at a grocery store, the event of catching a cold or some other virus from another shopper is independent from visit to visit and has a constant probability over the year, equal to 0.01. (a) In 100 trips to this store over the course of a year, the probability of catching a virus while shopping there is 100(0.01) = 1.0. What is wrong with this reasoning? (b) Find the correct probability in (a).*

**Solution**

Add comments to the solution.



### Ex 2.16
(Luca)
*Each day a hospital records the number of people who come to the emergency room for treatment.(a) In the first week, the observations from Sunday to Saturday are 10, 8, 14, 7, 21, 44, 60. Do you think that the Poisson distribution might describe the random variability of this phenomenon adequately. Why or why not?*

**Solution**
To assess whether the Poisson distribution might adequately describe the random variability of emergency room visits, we'll examine the data and compare it to properties of the Poisson distribution.


#### a)
```{r poisson_analysis, echo=TRUE}
# Data
er_visits <- c(10, 8, 14, 7, 21, 44, 60)
days <- c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")

# Basic statistics
mean_visits <- mean(er_visits)
var_visits <- var(er_visits)

# Print results
cat("Mean of visits:", round(mean_visits, 2), "\n")
cat("Variance of visits:", round(var_visits, 2), "\n")

# Plot the data
barplot(er_visits, names.arg = days, main = "Emergency Room Visits by Day",
        ylab = "Number of Visits", col = "skyblue")

# Add a line for the mean
abline(h = mean_visits, col = "red", lwd = 2)
legend("topleft", legend = "Mean", col = "red", lwd = 2)
```


1. The Poisson distribution has the property that its mean and variance are equal. In our data:
   - Mean: `r mean_visits`
   - Variance: `r var_visits`
   The large difference between these values suggests that the Poisson distribution may not be appropriate.

2. The bar plot shows a clear increasing trend throughout the week, with a sharp increase on Friday and Saturday. This pattern is not consistent with the Poisson distribution, which assumes a constant rate of events.

3. The Poisson distribution assumes:
   - Events occur independently
   - The average rate of occurrences is constant

   In this case, the number of ER visits doesn't satisfy these assumptions:
   - There may be dependencies (e.g., a local event affecting multiple people)
   - The rate clearly varies by day of the week

4. The data shows overdispersion (variance much larger than the mean), which is not characteristic of the Poisson distribution.

Given these observations, we can conclude that the Poisson distribution does not adequately describe the random variability of emergency room visits in this hospital. A more complex model that accounts for day-of-week effects and overdispersion (such as a negative binomial distribution or a time series model) would likely be more appropriate.

#### b)

**Solution**

Yes, we would expect the Poisson distribution to better describe the number of weekly admissions to the hospital for a rare disease.

1. **Rare events**: The Poisson distribution is particularly well-suited for modeling rare events. A rare disease, by definition, occurs infrequently.

2. **Independence**: Admissions for a rare disease are more likely to be independent of each other, especially if the disease is not contagious.

3. **Constant rate**: The occurrence of a rare disease is less likely to be affected by day-of-week patterns or other cyclical factors that we observed in general ER admissions. 

4. **No simultaneous occurrences**: With rare diseases, the probability of two or more admissions occurring simultaneously is extremely low, which aligns with another assumption of the Poisson distribution.

5. **Lower variance**: Rare events typically have a lower variance, which is more likely to be closer to the mean. 

6. **Small numbers**: The Poisson distribution is often used to model count data when the counts are small, which is likely the case for weekly admissions of a rare disease.

To illustrate this point, we can simulate weekly admissions for a hypothetical rare disease:

```{r rare_disease_simulation, echo=TRUE}
set.seed(123)
weeks <- 52
lambda <- 1.5  # Average 1.5 admissions per week for the rare disease
rare_disease_admissions <- rpois(weeks, lambda)

# Basic statistics
mean_admissions <- mean(rare_disease_admissions)
var_admissions <- var(rare_disease_admissions)

# Print results
cat("Mean of admissions:", round(mean_admissions, 2), "\n")
cat("Variance of admissions:", round(var_admissions, 2), "\n")

# Plot the data
hist(rare_disease_admissions, breaks = seq(-0.5, max(rare_disease_admissions) + 0.5, by = 1),
     main = "Histogram of Weekly Rare Disease Admissions",
     xlab = "Number of Admissions", ylab = "Frequency")

# Overlay Poisson distribution
x <- 0:max(rare_disease_admissions)
lines(x, dpois(x, lambda) * weeks, col = "red", type = "b", pch = 16)
legend("topright", legend = c("Observed", "Theoretical Poisson"),
       col = c("black", "red"), lty = 1, pch = c(22, 16))
```

**Comments on Solution:**

1. The mean (`r round(mean_admissions, 2)`) and variance (`r round(var_admissions, 2)`) of the simulated data are much closer to each other, which is characteristic of the Poisson distribution.

2. The histogram of simulated admissions closely follows the theoretical Poisson distribution (red line), indicating a good fit.

3. The number of admissions per week is small and varies within a narrow range, which is typical for rare events and well-described by the Poisson distribution.


### Ex 2.21
(Luca)
*Plot the gamma distribution by fixing the shape parameter k = 3 and setting the scale parameter $\theta$ = 0.5, 1, 2, 3, 4, 5. What is the effect of increasing the scale parameter? (See also Exercise 2.48.)*

**Solution**

To visualize the effect of increasing the scale parameter on the gamma distribution, we'll create a plot showing multiple gamma distributions with a fixed shape parameter and varying scale parameters.

```{r gamma_distribution_plot, echo=TRUE, fig.width=10, fig.height=6}
# Set parameters
k <- 3  # Shape parameter
theta <- c(0.5, 1, 2, 3, 4, 5)  # Scale parameters
colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")

# Create x values
x <- seq(0, 30, length.out = 1000)

# Plot
plot(x, dgamma(x, shape = k, scale = theta[1]), type = "l", col = colors[1],
     main = "Gamma Distribution with k = 3 and Varying theta",
     xlab = "x", ylab = "Density", ylim = c(0, 0.5))

# Add lines for other scale parameters
for (i in 2:length(theta)) {
  lines(x, dgamma(x, shape = k, scale = theta[i]), col = colors[i])
}

# Add legend
legend("topright", legend = paste("theta =", theta), col = colors, lty = 1, cex = 0.8)
```

**Comments on the solution:**

1. **Shape of the distribution**: As we increase the scale parameter $\theta$, we observe:

   a. The peak of the distribution shifts to the right (towards larger x values).
   b. The height of the peak decreases.
   c. The distribution becomes wider.

2. **Interpretation**: 
   - A larger scale parameter $\theta$ indicates greater variability and a shift in the distribution.
   - This can be explained by the relationship between the parameters of the gamma distribution:
     * The mean of the gamma distribution is $\mu = k\theta$
     * The variance is $\sigma^2 = k\theta^2$
   - As $\theta$ increases:
     * The mean increases linearly ($k\theta$)
     * The variance increases quadratically ($k\theta^2$)
   - This quadratic increase in variance relative to the mean explains the greater spread and variability we observe with larger $\theta$ values.


### Ex 2.26
Refer to Table 2.4 cross classifying happiness with family income.

**Solution**

Add comments to the solution.


### Ex 2.52
The pdf $f$ of a $N(\mu, \sigma^2)$ distribution can be derived from the standard normal pdf $\phi$ shown in equation $(2.9)$. (a) Show that the normal cdf $F$ relates to the standard normal cdf $\Phi$ by $F(y) = \Phi[(y - \mu)/\sigma]$. (b) From (a), show that $f(y) = (1/\sigma)\phi[(y - \mu)/\sigma]$, and show this is equation $(2.8)$.

**Solution**

Add comments to the solution.



### Ex 2.53
*If $Y$ is a standard normal random variable, with cdf $\Phi$, what is the probability distribution of $X = \Phi(Y)$? Illustrate by randomly generating a million standard normal random variables, applying the cdf function $\Phi()$ to each, and plotting histograms of the (a) $y$ values, (b) $x$ values.*

**Solution**

Add comments to the solution.


### Ex 2.70
*The beta distribution is a probability distribution over (0, 1) that is often used in applications for which the random variable is a proportion. The beta pdf is ...*

**Solution**

Add comments to the solution.



## FSDS - Chapter 3 

### Ex 3.18
*Sunshine City, which attracts primarily retired people, has 90,000 residents with a mean age of 72 years and a standard deviation of 12 years. The age distribution is skewed to the left. A random sample of 100 residents of Sunshine City has y = 70 and s = 11...*
**Solution**

Add comments to the solution.


### Ex 3.28
*A survey is planned to estimate the population proportion $\pi$ supporting more government action to address global warming. For a simple random sample, if $\pi$ may be near 0.50, how large should n be so that the standard error of the sample proportion is 0.04?*
**Solution**

Add comments to the solution.


### Ex 3.24 (use R)
*Construct a population distribution that is plausible for Y = number of alcoholic drinks in the past day. Use the following steps:*

**Solution**

Add comments to the solution.


## FSDS - Chapter 4 

### Ex 4.14
*Using the Students data file, for the corresponding population, construct a 95% confidence interval (a) for the mean weekly number of hours spent watching TV; (b) to compare females and males on the mean weekly number of hours spent watching TV. In each case, state assumptions, including the practical importance of each, and interpret results.*

**Solution**

Add comments to the solution.


### Ex 4.16
*The Substance data file at the book’s website shows a contingency table formed from a survey that asked a sample of high school students whether they have ever used alcohol, cigarettes, and marijuana. Construct a 95% Wald confidence interval to compare those who have used or not used alcohol on whether they have used marijuana, using (a) formula (4.13); (b) software. State assumptions for your analysis, and interpret results.*

**Solution**

Add comments to the solution.



### Ex 4.48
*For a simple random sample of $n$ subjects, explain why it is about 95% likely that the sample proportion has error no more than $1/\sqrt{n}$ in estimating the population proportion. (Hint: To show this "$1/\sqrt{n}$ rule," find two standard errors when $\pi = 0.50$, and explain how this compares to two standard errors at other values of $\pi$.) Using this result, show that $n = 1/M^2$ is a safe sample size for estimating a proportion to within $M$ with 95% confidence.*

**Solution**

Add comments to the solution.



## FSDS - Chapter 5 

### Ex 5.2
*When a government does not have enough money to pay for the services that it provides, it can raise taxes or it can reduce services. When the Florida Poll asked a random sample of 1200 Floridians which they preferred, 52% (624 of the 1200) chose raise taxes and 48% chose reduce services. Let $\pi$ denote the population proportion of Floridians who would choose raising taxes. Analyze whether this is a minority of the population ($\pi$ < 0.50) or a majority ($\pi$ > 0.50) by testing H0 : $\pi$ = 0.50 against Ha : $\pi$ $\neq$ 0.50. Interpret the P -value. Is it appropriate to “accept H0 ? Why or why not?*
**Solution**

Add comments to the solution.



### Ex 5.12
*The example in Section 3.1.4 described an experiment to estimate the mean sales with a proposed menu for a new restaurant. In a revised experiment to compare two menus, on Tuesday of the opening week the owner gives customers menu A and on Wednesday she gives them menu B. The bills average $22.30 for the 43 customers on Tuesday (s = 6.88) and $25.91 for the 50 customers on Wednesday (s = 8.01). Under the strong assumption that her customers each night are comparable to a random sample from the conceptual population of potential customers, show how to compare the mean sales for the two menus based on (a) the P -value of a significance test, (b) a 95% confidence interval. Which is more informative, and why? (When used in an experiment to compare two treatments to determine which works better, a two-sample test is often called an A/B test.).*

**Solution**

Add comments to the solution.



### Ex 5.50
*A random sample of size 40 has y = 120. The P -value for testing H0: $\mu$ = 100 against H : $\mu$ $\neq$ 100 is 0.057. Explain what is incorrect about each of the following interpretations of this P-value, and provide a proper interpretation.*
**Solution**

Add comments to the solution.
